---
title: "Group Project: Kentucky counties — Incarceration vs Poverty"
author: "Dr. Mike LeVan"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
params:
  incarceration_csv: "kentucky_county_incarceration_2019_2024.csv"
  poverty_csv: "kentucky_county_poverty_2019_2024.csv"
  years: [2019,2020,2021,2022,2023,2024]
---

# Group Project: Kentucky counties — Incarceration vs Poverty (2019-2024). {.unnumbered}

## Data

This group project uses two prepared CSVs:

- kentucky_county_incarceration_2019_2024.csv — county-year incarceration counts and rates (per 100k).
- kentucky_county_poverty_2019_2024.csv — county-year poverty counts and poverty rate (%).

These can be found on our class Canvas page (Oct 24).


## Goals

- Use the two datasets and explore whether counties in Kentucky with higher poverty rates have higher incarceration rates.
- Use descriptive statistics, scatterplots and boxplots, correlation, a simple linear regression, and make predictions of incarceration rate from poverty.

## Instructions

1) Read and inspect the data. Read the incarceration data into a variable named `incar` 
and the poverty data into a variable named `pov`.

```{r read-data}
#| eval: false
#| include: false
incar <- read_csv(incarceration_csv, show_col_types = FALSE)
pov   <- read_csv(poverty_csv, show_col_types = FALSE)

# Inspect columns and a few rows
glimpse(incar)
glimpse(pov)

head(incar, 3)
head(pov, 3)
```

  - Describe the key columns in each dataset. Which variables would be best to use 
in this report? Why? Discuss any issues that might be present.


2) Basic descriptive statistics
- For the two datasets (all years) and by year, report:
  - Number of counties (rows)
  - Mean, median, standard deviation of incarceration_rate_per_100k and poverty_rate.
  Use the variable names mean_incar, med_incar, sd_incar for incarceration rate and
  mean_pov, med_pov, sd_pov for poverty rate.
  - Use dplyr functions: summarise(), group_by(), mean(), median(), sd().
  - Discuss you findings and create a table to show these results.
```{r descriptives}
#| eval: false
#| include: false
# Overall
df %>%
  summarise(
    n = n(),
    mean_incar = mean(incarceration_rate_per_100k, na.rm = TRUE),
    med_incar = median(incarceration_rate_per_100k, na.rm = TRUE),
    sd_incar = sd(incarceration_rate_per_100k, na.rm = TRUE),
    mean_pov = mean(poverty_rate, na.rm = TRUE),
    med_pov = median(poverty_rate, na.rm = TRUE),
    sd_pov = sd(poverty_rate, na.rm = TRUE)
  ) %>% print()

# By year
df %>%
  group_by(year) %>%
  summarise(
    n = n(),
    mean_incar = mean(incarceration_rate_per_100k, na.rm = TRUE),
    mean_pov = mean(poverty_rate, na.rm = TRUE)
  ) %>%
  arrange(year) %>%
  print(n = Inf)
```

3) Boxplots (poverty rate by year)
- Create boxplots of poverty_rate by year to see how poverty rates vary across years.
- Use a different color fill for the boxes.
- Discuss any trends or patterns you observe, if they data shows any.
```{r boxplot-poverty, fig.height=5, fig.width=7}
#| eval: false
#| include: false
ggplot(df, aes(x = factor(year), y = poverty_rate)) +
  geom_boxplot(fill = year) +
  labs(x = "Year", y = "Poverty rate (%)", title = "P
overty Rate by Year (KY counties)") +
  theme_minimal()
```

4) Correlation
- Compute Pearson correlation between poverty_rate and incarceration_rate_per_100k overall and by year.
```{r correlation}
#| eval: false
#| include: false
# overall Pearson
cor_overall <- cor.test(df$poverty_rate, df$incarceration_rate_per_100k, method = "pearson")
cor_overall$estimate
cor_overall$p.value

# by year
df %>%
  group_by(year) %>%
  summarise(
    r = cor(poverty_rate, incarceration_rate_per_100k, use = "complete.obs"),
    n = sum(!is.na(poverty_rate) & !is.na(incarceration_rate_per_100k))
  ) %>%
  print(n = Inf)
```

5) Scatterplots (poverty vs incarceration)
- Make a scatterplot of poverty_rate (x) vs incarceration_rate_per_100k (y).
- Describe the scatterplot. Is there a visible relationship? Is it strong or weak?

```{r scatter-all, fig.height=5, fig.width=7}
#| eval: false
#| include: false
ggplot(df, aes(x = poverty_rate, y = incarceration_rate_per_100k, color = factor(year))) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +   # overall regression line
  labs(x = "Poverty rate (%)", y = "Incarceration rate (per 100k)",
       color = "Year", title = "Poverty vs Incarceration (KY counties)") +
  theme_minimal()
```

7) Linear regression
- Fit a model to predict incarceration_rate_per_100k from poverty_rate using lm().
- Use the `summary()` function to report coefficients and the R-squared value. 
Explain what these values mean in plain language.

```{r simple-lm}
#| eval: false
#| include: false
simple_model <- lm(incarceration_rate_per_100k ~ poverty_rate, data = df)
summary(simple_model)

# tidy coefficients
tidy(simple_model, conf.int = TRUE)
```

Interpretation guidance (write this in your report)
- Report the slope and intercept. In plain language: "The slope means that for every 1 percentage-point increase in poverty rate, the incarceration rate increases/decreases by X per 100k (on average)."
- Report R-squared: how much variation the model explains.
- Note: this is association, not causation.

Overlay the regression line on the scatter (already shown above with geom_smooth).

8) Make predictions
- Predict incarceration rate for hypothetical poverty values (e.g., 10%, 15%, 25%).
```{r predict-examples}
#| eval: false
#| include: false
new_pov <- tibble(poverty_rate = c(10, 15, 25))
predict(simple_model, newdata = new_pov, interval = "prediction", level = 0.95)
```

- Interpret the prediction intervals: they show uncertainty around a predicted county's incarceration rate.


## Grading  

  - Code runs and outputs saved: 30%
  - Visualizations and descriptive stats: 30%
  - Correct correlation, regression, and interpretation: 30%
  - Clarity of write-up and ethical framing: 10%

## Ethics & limitations (short to include in write-up)
  - This analysis shows association, not causation. Do not claim poverty causes incarceration based on this simple model.
  - Data limitations: county-level rates can be unstable in small counties; reporting differences and missing data may affect results.
  - Race and demographic variables (if used) are descriptive of population composition; interpreting them requires careful consideration of structural and historical context.


Good luck — work together, document your steps, and keep your interpretations cautious and respectful.